---
title: "Datacamp - Communicating with Data in the Tidyverse"
author: "A. Rasheed Narejo"
date: "06-07-2018"
output:
  html_document:
    css: styles.css
    df_print: kable
    highlight: monochrome
    number_sections: yes
    theme: cosmo
    toc: yes
    toc_depth: 4
    toc_float: no
  pdf_document:
    latex_engine: xelatex
    toc: yes
    toc_depth: '4'
subtitle: Reduction in working hours and increase in hourly wages in Europe between
  1996 and 2006
---

## Summary
The objective of this assignment is to use tidyverse package and Rmarkdown to create professional quality graphs and documents.

The **International Labour Organization** (ILO) has many data sets on working conditions. For example, one can look at how weekly working hours have been decreasing in many countries of the world, while monetary compensation has risen. In this report, the reduction in weekly working hours in European countries is analysed, and a comparison between 1996 and 2006 is made. All analysed countries have seen a decrease in weekly working hours since 1996 – some more than others.

## visualization of labour market data
This is a personal project to replicate the data manipulation/visualization in DataCamp course [Communicating with Data in the Tidyverse](https://www.datacamp.com/courses/communicating-with-data-in-the-tidyverse).

## load required libraries
The project requires tidyverse packages for data manipulation(dplyr) and visualization (ggplot2).

```{r message=FALSE, warning=TRUE}
library(tidyverse)
library(ggthemes)
library(forcats)
```

## Data access
The ilo labour market files have been downloaded and saved in local data folder. We can use also use `list.files("data")` to check the files details.

### Load data
```{r}
# load files
load("data/ilo_hourly_compensation.RData")
load("data/ilo_working_hours.RData")
```

### Merge data in one file
Since, the **hourly wages** and **working hours per week** are provided in two separate files, we will first merge two files together using inner join function on country and year to create a single data frame.

```{r message=FALSE}
ilo_data <- ilo_hourly_compensation %>%
    inner_join(ilo_working_hours, on = c(country, year))
```

### Convert year and country as factor variables
```{r}
ilo_data$year <- as.numeric(ilo_data$year)

ilo_data_factored <- ilo_data %>%
  mutate(year = as.factor(year),
        country = as.factor(country))
```

As data exploration, we can check first few rows of data to get an idea of it. We can use knitr package to beautify the tables.

```{r}
ilo_data_factored %>%
    head() %>%
      knitr::kable()
```

## Filter data for european countries  and year 1981 and 2006
Our primary objective for this assignment is to check the weekly hours worked and average hourly wage trend in European countries.
```{r}
# vector of all european countries
european_countries <- c("Finland", "France", "Italy", "Norway", "Spain", "Sweden", "Switzerland", "United Kingdom", "Belgium", "Ireland", "Luxembourg", "Portugal", "Netherlands", "Germany", "Hungary", "Austria", "Czech Rep.")
```

As a first step, lets check trend of average working hours and average hourly wages across europe over the years.

```{r}
ilo_data_factored %>%
  filter(country %in% european_countries) %>%
  group_by(year) %>%
  summarise(avg_hourly_compensation = mean(hourly_compensation),
            avg_working_hours = mean(working_hours)) %>%
    ggplot(aes(x = avg_hourly_compensation, y = avg_working_hours)) + 
      geom_point() +
      geom_smooth(method='lm') + 
      labs(x = "Average hourly compensation", 
           y = "Average working hours", 
           title = "ILO data comparison of working hours and average hourly rate in european countries",
           subtitle = "Average for each year between 1980 and 2008",
           caption = "source: ILO")
```

There is a definitive trend of increasing hourly wages and declining average weekly working hours over the years.

```{r}
ilo_data_final <- ilo_data_factored %>%
    filter(country %in% european_countries) %>%
    filter(year == 1996 | year == 2006)
```

```{r generating_summary_statistics, echo = TRUE}
# Some summary statistics
ilo_data_final %>%
  group_by(year) %>%
  summarize(mean_hourly_compensation = mean(hourly_compensation),
            mean_working_hours = mean(working_hours)) %>% 
  knitr::kable()
```


## Plotting data using ggplot
### Build a base plot
```{r}
base_plot <- ggplot(ilo_data_final, aes(x = hourly_compensation, y = working_hours) ) + 
  geom_point() + 
  labs(x = "Compensation per hour",
       y = "Hour per week",
       title = "Lesser working hours lead to higher wages",
       subtitle = "Working hours and hourly compensation in European countries, 1980 vs. 2006",
       caption = " Data source, ILO") + 
  facet_grid(. ~ year)
base_plot
```

### Add minimal theme on base plot
```{r}
base_plot + theme_minimal()
```

### applying economist theme from ggthemes
```{r}
base_plot + theme_economist()
```

### customise mininal base theme
```{r}
base_plot + theme_minimal() + 
    theme(
    text = element_text(family = "Arial", color = "gray25"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(color = "gray30"),
    plot.background = element_rect(fill = "gray95"),
    plot.margin = unit(c(5, 10, 5, 10), units = "mm")
  )
```

### define own custom theme and apply on base_plot
```{r}
theme_ilo <- function() {
  theme_minimal() +
  theme(
    text = element_text(family = "Arial", color = "gray25"),
    plot.subtitle = element_text(size = 12),
    plot.caption = element_text(color = "gray30"),
    plot.background = element_rect(fill = "gray95"),
    plot.margin = unit(c(5, 10, 5, 10), units = "mm")
  )

}

ilo_plot <- base_plot + theme_ilo()
ilo_plot
```

### Further customisation
```{r}
ilo_plot + theme(
    # Change the background fill and color
    strip.background = element_rect(fill = "gray60",  color = "gray95"),
    # Change the color of the text
    strip.text = element_text(color = "white")
  )


```

## Creating dotplot

### Create a Basic dot Plot with arrows and labels
```{r}
ggplot(ilo_data_final, aes(x = working_hours, y = country)) + 
  geom_path(
# Add an arrow to each path
  arrow = arrow(length = unit(1.5, "mm"), type = "closed")) + 
    # Add a geom_text() geometry
  geom_text(aes(x = working_hours,
              y = country,
              label = round(working_hours, 1))
        )
```
### ordering dot plots by working hours
```{r}
# Reorder country factor levels
ilo_data_final <- ilo_data_final %>%
  # Arrange data frame
  arrange(year) %>%
  # Reorder countries by working hours in 2006
  mutate(country = fct_reorder(country,
                               working_hours,
                               last))

# Plot again
ggplot(ilo_data_final) +
  geom_path(aes(x = working_hours, y = country),
            arrow = arrow(length = unit(1.5, "mm"), type = "closed")) +
    geom_text(
          aes(x = working_hours,
              y = country,
              label = round(working_hours, 1))
          )

```


```{r}
# Save plot into an object for reuse
ilo_dot_plot <- ggplot(ilo_data_final) +
  geom_path(aes(x = working_hours, y = country),
            arrow = arrow(length = unit(1.5, "mm"), type = "closed")) +

    # Specify the hjust aesthetic with a conditional value
    geom_text(
          aes(x = working_hours,
              y = country,
              label = round(working_hours, 1),
              hjust = ifelse(year == "2006", 1.5, -0.3)
            ),
          # Change the appearance of the text
          size = 3,
          family = "Arial",
          color = "gray25"
          )

ilo_dot_plot

```

### Customising plot for different audience and devices
```{r}
ilo_dot_plot <- ilo_dot_plot +
  # Add labels to the plot
  labs(
    x = "Working hours per week",
    y = "Country",
    title = "People work less in 2006 compared to 1996",
    subtitle = "Working hours in European countries, development since 1996",
    caption = "Data source: ILO, 2017"
  ) +
  # Apply your theme
  theme_ilo() +
  # Change the viewport
  coord_cartesian(xlim = c(25, 41))
  
# View the plot
ilo_dot_plot
```








